Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-08-01 experiments.tex
Final fixes, removed accepted commennts

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we use \comment{AA:da ``we will be using'' a ``we use''} a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The \comment{AA:Ho eliminato ``same'' che non avevo capito a cosa si riferisse} philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures require designers
to trade between power \comment{AA:Prima era ``let designers easily trade'', ma mi suonava strano}, energy and performances, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

\subsection{Parameter Space}
In general, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. 
The first is directly related to the VLIW core and includes the size
of the register files and the number of functional
units of each type\comment{AA:Da ``for each type'' a ``of each type''}. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal 
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
do\_modulo\_scheduling \comment{AA:Non ho capito; E' un nuovo item questo?} Modular scheduling 
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm
(MOGA)~\cite{knowles_techrep06}. To the best of our knowledge \comment{AA:da ``At best ...'' a ``To the best'': lo leggo quasi sempre cosi}, genetic
approaches in general still represent the best compromise between
efficiency (i.e. time required for exploration) and accuracy of the
reported Pareto Sets.  Note that, for the reasons described above,
mono-objective approaches have been explicitly excluded from this
analysis.  We also exclude other approaches (e.g.
dependency based~\cite{givargis_tvlsi02}) that make use of some
a-priori knowledge about the role and meaning of each parameter, since
our aim is evaluate the capacity of focusing on interesting parameter
regions without any external judgement or help based on heuristics. \comment{AA: Nella frase precedente sono passato da da ``Same decision has been taken for'' a ``We also exclude'' perche a un lettore disattento potrebbe sembrare che volevamo dire che anche givargis and co. avessero adottato la nostra decision di escludere genetic approaches}
Finally, it should be pointed out that in this introductory work we
will explicitly use the \emph{SPEA}~\cite{zitzler_eurogen01} variant of MOGA,
remanding to works like~\cite{zitzler_ec00}~\cite{zitzler_tec03} for a
detailed analysis on how it differs \comment{AA: from what?} and what could be expected from
similar multi-objective strategies.

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly dependent upon the application being considered.
These results are obtained in about 5 eras and demonstrate that with a so limited budget PS\comment{AA: ho tolto ``still'' perche' non ne ho capito il senso} can compete with
genetic approach while even discovering very different Pareto fronts. \comment{AA: nella frase precedente cambiato da ``, and the main aim was demonstrating'' a "and demonstrate"}
Considering the extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective.
A comparison between the variation ranges of both power dissipation and execution time, obtained by PS and GA for different benchmarks is shown in Fig.~\ref{fig:dispersion}(a) and (b) respectively. \comment{La frase precedente e' un rifrasamento (esiste sto termine in italiano?) di ``A
comparison between the variation range for different benchmarks
between a PS and a GA exploration for both power dissipation and
execution time is shown in Fig.~\ref{fig:dispersion}(a) and (b)
respectively.''}
 As it can be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than the one provided by a
GA exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

@@

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA \comment{AA: ho tolto ``still'' perche' non ne ho capito il senso} shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the \comment{AA: cambiato da ``availability '' a ``existence''} of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
is worth pointing out how a definitive claim about the superiority of
the PS behaviour is strictly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevertheless, the proposed algorithm introduces a new
perspective on \comment{AA: ho rimosso `` thinking'' perche' altrimenti avrei preferito ``a new way of thinking''} the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software design space exploration.

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

\subsection{Parameter Space}
In general, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. 
The first is directly related to the VLIW core and includes the size
of the register files and the number of functional
units of each type. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal and loop unrolling
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
\item \emph{do\_modulo\_scheduling}: Modular scheduling of loop code
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm
(MOGA)~\cite{knowles_techrep06}. At the best of our knowledge, genetic
approaches in general still represent the best compromise between
efficiency (i.e. time required for exploration) and accuracy of the
reported Pareto Sets.  Note that, for the reasons described above,
mono-objective approaches have been explicitly excluded from this
analysis.  We also exclude other approaches (e.g.  dependency
based~\cite{givargis_tvlsi02}) that make use of some a-priori
knowledge about the role and meaning of each parameter, since our aim
is evaluate the capacity of focusing on interesting parameter regions
without any external judgement or help based on heuristics.  Finally,
it should be pointed out that in this introductory work we will
explicitly use the \emph{SPEA}~\cite{zitzler_eurogen01} variant of
MOGA, remanding to works like~\cite{zitzler_ec00}~\cite{zitzler_tec03}
for a detailed analysis on how it differs from other genetic approaches
and what could be expected from similar multi-objective strategies.

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly dependent upon the application being considered.
These results are obtained in about 5 eras and demonstrate that with a so limited budget PS can compete with
genetic approach while even discovering very different Pareto fronts. 
Considering the extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective.
A comparison between the variation ranges of both power dissipation
and execution time, obtained by PS and GA for different benchmarks is
shown in Fig.~\ref{fig:dispersion}(a) and (b) respectively.  As it can
be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than the one provided by a GA
exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

@@

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
is worth pointing out how a definitive claim about the superiority of
the PS behaviour is strictly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevertheless, the proposed algorithm introduces a new
perspective on the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software design space exploration.

### 2014-08-01 paper.tex
Final fixes, removed accepted commennts

\section{Conclusions}
In this work we presented PS, a multiobjective strategy which 
introduces the concept of Parameter Space representation of Pareto
Front. A case study of a VLIW architecture involving strong
hardware/software dependencies has been analyzed \comment{AA: da ``carried out'' a ``analyzed'', carry out a case study (effettuare un caso studio) mi suonava strano } carried out to evaluate
effectiveness with different amount of simulations budget.
A qualitative/quantitative comparison against a widespread multiobjective genetic
approach showed how the proposed strategy can result in a different
distribution of Pareto points which trades solutions granularity with
improved Pareto extension.

>>>>

\section{Conclusions}
In this work we presented PS, a multiobjective strategy which 
introduces the concept of Parameter Space representation of Pareto
Front. A case study of a VLIW architecture involving strong
hardware/software dependencies has been analyzed to evaluate
effectiveness with different amount of simulations budget.
A qualitative/quantitative comparison against a widespread multiobjective genetic
approach showed how the proposed strategy can result in a different
distribution of Pareto points which trades solutions granularity with
improved Pareto extension.
